A Bayesian model of latitudinal temperature gradients applied to the Early Eocene

### 2023-02-12 manuscript/manuscript.Rmd
discussion

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).

These high-t proxies at high latitudes, combined with the overall high variability of the proxy record, is also the cause f...

>>>>

In latitudes beyond the polar circle, the discrepancy between our model predictions and those of earlier, proxy-based models increases, as those predict almost linearly decreasing SSTs towards the poles, whereas our median model prediction suggests only a slight decrease beyond the polar circle. However, the scarcity of temperature records in this range leads to widening credible intervals, including the possibility of stronger temperature decreases. Polar temperature estimates from our model are thus conservative in that they admit large uncertainty where data is absent, which is desirable. However, the presence of high proxy-derived temperature estimates at ~ 60 deg latitudes forces the modeled median temperature curve to be too high at ~ 24 degC, relative to the temperatures indicated by the high-latitude mangrove communities (15.6 - 22.5 degC).

The very high variability of the proxy record in mid-latitudes results in large uncertainties on the shape of temperature gradient and on the GMSsT, and demands closer scrutiny. Biases and errors in the proxy reconstructions likely contribute to the observed variability. A host of factors influences reconstructed temperatures from geochemical proxies [@Hollis2019]. We tried to mitigate biases by excluding d18O measurements from recrystallised fossils, and by using the screened data set of @Hollis2019, but systematic offsets remain between mostly warm temperatures derived from TEX86, and cooler temperatures derived from d18O and delta47. Temporal averaging also likely contribute to the large variability of the EECO proxy data. The EECO spans ~ 4 million years, and
